---
title: "Update Joint Modeling 07/10"
format:
  pdf:
    papersize: a4
    geometry: margin=2cm
    fig-width: 7
    fig-height: 4
editor: visual
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(here)
library(ggplot2)
library(dplyr)
library(tidyr)

knitr::opts_chunk$set(dev = "png", dpi = 300)

res_all <- readRDS(here::here(
  "results", "data", "1aad8238bd68a72a", "aggregation_summary_default.rds"
))$aggregation$summary

# Method x scenario rows where every replicate errored carry no information; they are dropped and named below.
failed <- res_all |> filter(prop_error == 1)
res <- res_all |> filter(prop_error < 1)

method_labels <- c(
  classical_ml = "Classical ML",
  multiple_imputation = "MI + closed form",
  reweighting = "Reweighting",
  LSPIM = "LSPIM"
)
method_colours <- c(
  "Classical ML" = "#0072B2",
  "MI + closed form" = "#E69F00",
  "Reweighting" = "#009E73",
  "LSPIM" = "#CC79A7"
)
dropout_labels <- c(half_missing = "Half-missing", three_obs_minimum = "Three observations minimum")

# Log scenarios: only the large (-2.42, 1.95) and small (-0.385, 0.31) crossing pairs and the null are reported.
is_pair <- function(b1, b3, p1, p3) abs(b1 - p1) < 1e-6 & abs(b3 - p3) < 0.005
res <- res |>
  mutate(
    log_effect = case_when(
      time_trend != "log" ~ NA_character_,
      beta1 == 0 & beta3 == 0 ~ "null",
      is_pair(beta1, beta3, -2.42, 1.948) ~ "large",
      is_pair(beta1, beta3, -0.385, 0.310) ~ "small",
      TRUE ~ "excluded"
    )
  ) |>
  filter(is.na(log_effect) | log_effect != "excluded") |>
  mutate(
    method = factor(method_labels[method], levels = method_labels),
    dropout = factor(dropout_labels[dropout_mechanism], levels = dropout_labels),
    n = factor(n),
    effect = case_when(
      time_trend == "linear" & beta3 == 0 ~ "null (beta3 = 0)",
      time_trend == "linear" ~ paste0("beta3 = ", signif(beta3, 3)),
      log_effect == "null" ~ "null (identical arms)",
      log_effect == "large" ~ "large (beta1 = -2.42, beta3 = 1.95)",
      log_effect == "small" ~ "small (beta1 = -0.385, beta3 = 0.31)"
    ),
    effect = factor(effect, levels = c(
      "beta3 = 0.22", "beta3 = 0.035", "null (beta3 = 0)",
      "large (beta1 = -2.42, beta3 = 1.95)", "small (beta1 = -0.385, beta3 = 0.31)", "null (identical arms)"
    )),
    # Short column labels for the tables
    eff_short = case_when(
      beta3 == 0 ~ "0",
      log_effect == "large" ~ "large",
      log_effect == "small" ~ "small",
      TRUE ~ as.character(signif(beta3, 3))
    ),
    eff_short = factor(eff_short, levels = c("0.22", "0.035", "large", "small", "0")),
    drop_short = ifelse(dropout_mechanism == "half_missing", "HM", "3O")
  )

lin <- res |> filter(time_trend == "linear")
logt <- res |> filter(time_trend == "log")

# Three significant digits, keeping trailing zeros; NA as a dash.
fmt3 <- function(x) ifelse(is.na(x), "-", formatC(signif(x, 3), format = "fg", digits = 3, flag = "#"))
fmt3 <- Vectorize(fmt3)
fmt3_trim <- function(x) sub("\\.$", "", fmt3(x))

# Wide table: rows n x method, columns dropout x effect.
metric_table <- function(df, metric, caption) {
  df |>
    mutate(value = fmt3_trim(.data[[metric]]), col = paste(drop_short, eff_short, sep = ": ")) |>
    arrange(drop_short != "HM", eff_short) |>
    select(n, method, col, value) |>
    pivot_wider(names_from = col, values_from = value) |>
    arrange(n, method) |>
    rename(Method = method) |>
    knitr::kable(caption = caption, align = "l")
}

metric_plot <- function(df, metric, ylab, hline = NULL, log_y = FALSE) {
  p <- ggplot(df, aes(
    x = n, y = .data[[metric]], colour = method, shape = effect, linetype = effect,
    group = interaction(method, effect)
  )) +
    geom_line() +
    geom_point(size = 2) +
    facet_wrap(~dropout) +
    scale_colour_manual(values = method_colours, drop = TRUE) +
    labs(x = "Number of clusters", y = ylab, colour = "Method", shape = "Effect", linetype = "Effect") +
    theme_minimal() +
    theme(legend.position = "bottom", legend.box = "vertical")
  if (n_distinct(df$effect) == 1) p <- p + guides(shape = "none", linetype = "none")
  if (!is.null(hline)) p <- p + geom_hline(yintercept = hline, linetype = "dotted")
  if (log_y) p <- p + scale_y_log10()
  p
}

status_levels <- c(
  prop_converged_ok = "Converged",
  prop_converged_warning = "Converged (warning)",
  prop_converged_singular = "Converged (singular)",
  prop_not_converged = "Not converged",
  prop_error = "Error"
)

# Convergence pooled over effect sizes: convergence barely depends on beta3.
convergence_long <- function(df) {
  df |>
    group_by(method, dropout, n) |>
    summarise(across(all_of(names(status_levels)), mean), .groups = "drop") |>
    pivot_longer(all_of(names(status_levels)), names_to = "status", values_to = "prop") |>
    mutate(status = factor(status_levels[status], levels = status_levels))
}

convergence_plot <- function(df) {
  ggplot(convergence_long(df), aes(x = n, y = prop, fill = status)) +
    geom_col(width = 0.75) +
    facet_grid(dropout ~ method, labeller = label_wrap_gen(18)) +
    scale_fill_manual(values = c("#009E73", "#56B4E9", "#F0E442", "#D55E00", "#999999"), drop = FALSE) +
    labs(x = "Number of clusters", y = "Proportion of replicates", fill = NULL) +
    theme_minimal() +
    theme(legend.position = "bottom", axis.text.x = element_text(size = 7))
}

convergence_table <- function(df, caption) {
  df |>
    group_by(method, drop_short, n) |>
    summarise(
      singular = mean(prop_converged_singular),
      not_conv = mean(prop_not_converged),
      .groups = "drop"
    ) |>
    pivot_longer(c(singular, not_conv), names_to = "what", values_to = "value") |>
    mutate(col = paste(drop_short, ifelse(what == "singular", "singular", "not conv."), sep = ": "), value = fmt3_trim(value)) |>
    arrange(drop_short != "HM") |>
    select(n, method, col, value) |>
    pivot_wider(names_from = col, values_from = value) |>
    arrange(n, method) |>
    rename(Method = method) |>
    knitr::kable(caption = caption, align = "l")
}

lambda_table <- function(df, caption) {
  df |>
    filter(method == "MI + closed form") |>
    mutate(value = fmt3_trim(mean_mi_lambda_beta3), col = paste(drop_short, eff_short, sep = ": ")) |>
    arrange(drop_short != "HM", eff_short) |>
    select(n, col, value) |>
    pivot_wider(names_from = col, values_from = value) |>
    arrange(n) |>
    knitr::kable(caption = caption, align = "l")
}
```

# Overview methods

| Method | Configuration | "Not converged" when |
|----|----|----|
| Classical ML | `lme4::lmer`, ML (REML = FALSE); Wald $z$ test on $\beta_3$ | optimizer code $\neq 0$ or an lme4 convergence message |
| MI + closed form | `mice` `2l.pmm`, 3 imputations, `maxit` = 10; closed form (CbC) per imputation, pooled with Rubin's rules; Wald $z$ test | never, if all fits succeed (any failed imputation fit = error) |
| Closed form + iterative reweighting | Reweighting until max $\Delta\beta < 10^{-6}$, at most 30 iterations; Wald $z$ test | the loop hits 30 iterations |
| LSPIM | Pseudo-score GEE model, Holm-adjusted interaction test; only run for $n \leq 50$ | any of its GEEs does not converge |

-   All tests at $\alpha = 0.05$; coverage of the 95% Wald interval.
-   Singular fits (boundary random-effects covariance, or a repaired $D$ for the closed-form methods) are kept in all metrics and shown as a separate status.
-   Only replicates that end in an error are excluded from the metrics.

```{r}
#| label: failed-note
#| results: asis

if (nrow(failed) > 0) {
  cat(
    "::: callout-warning\n## Missing results\n\n",
    "All replicates errored for: ",
    paste(unique(method_labels[failed$method]), collapse = ", "),
    " (", nrow(failed), " scenarios). ",
    "These rows are left out of all tables and plots.\n\n",
    "LSPIM: package `geessbin` was not installed on the computer that ran the simulation; a rerun is pending.\n:::\n",
    sep = ""
  )
}
```

\newpage

# Simulation setup

Linear model: $$ y_{ij} = \beta_0 + \beta_1 Treatment_i + \beta_2 Time_j + \beta_3 Treatment_i \times Time_j + b_{0i} + b_{1i} Time_j + \varepsilon_{ij}, \quad Time_j = j = 0, \dots, 11 $$

Log model: the same, with $Time_j$ replaced by $\log(1 + j)$ in both the mean and the random slope.

| Effect | Parameter | Linear | Log |
|----|----|----|----|
| Intercept | $\beta_0$ | 2.4562 | 2.4562 |
| Treatment effect | $\beta_1$ | 0 | $-2.42$ (large), $-0.385$ (small), 0 (null) |
| Time effect | $\beta_2$ | 0.2792 | 1.2359 |
| Treatment $\times$ time | $\beta_3$ | 0.22, 0.035, 0 (null) | 1.948 (large), 0.310 (small), 0 (null) |
| Random intercept variance | $d_{11}$ | 7.3174 | 7.3174 |
| Random slope variance | $d_{22}$ | 0.2239 | 4.3875 |
| Random effects covariance | $d_{12}$ | $-0.4985$ | $-2.2067$ |
| Residual variance | $\sigma^2$ | 3.1508 | 3.1508 |

-   Linear parameters taken from the toenail data ($\beta_3 = 0.035$); $\beta_3 = 0.22$ added as a larger effect.
-   Log parameters: $\beta_2$, $d_{22}$ and $d_{12}$ rescaled by $11 / \log 12$, so the rise and the slope variance at $t = 11$ match the linear scenario.
-   Log crossing: arms cross at $t \approx 2.5$; the treatment difference at $t = 11$ equals that of the linear scenario with $\beta_3 = 0.22$ (large) or $0.035$ (small).
-   Log null: identical arms ($\beta_1 = \beta_3 = 0$).

| Variable | Value |
|----|----|
| Number of clusters | 10, 20, 50, 100, 1000 |
| Number of repeated measures | 12 |
| Missingness pattern | "Half-missing" (HM): half of the clusters have no missingness, the other half have a uniform probability to keep 1--11 observations; dropout is monotone. |
| | "Three observations minimum" (3O): as half-missing, but with at least three observations per cluster. |
| Number of simulations per scenario | 5000 |

\newpage

# Results - linear time trend

## Mean squared error of $\beta_3$

```{r}
#| label: tbl-lin-mse
metric_table(lin, "mse_beta3", "MSE of beta3 (HM = half-missing, 3O = three observations minimum; columns = true beta3).")
```

```{r}
#| label: fig-lin-mse
#| fig-cap: "MSE of beta3 (log scale)."
metric_plot(lin, "mse_beta3", "MSE (beta3)", log_y = TRUE)
```

-   MSE falls with the number of clusters at the same rate for all methods; the effect size and dropout mechanism hardly matter.
-   MI has the lowest MSE at small $n$; from $n = 50$ onwards classical ML and reweighting are slightly better and similar to each other.

\newpage

## Type I error

```{r}
#| label: tbl-lin-t1e
metric_table(filter(lin, beta3 == 0), "type1_error", "Type I error rate of the interaction test (beta3 = 0).")
```

```{r}
#| label: fig-lin-t1e
#| fig-cap: "Type I error rate (dotted line: nominal 0.05)."
metric_plot(filter(lin, beta3 == 0), "type1_error", "Type I error rate", hline = 0.05)
```

-   All methods are anti-conservative at small $n$; classical ML is inflated the most, MI and reweighting are similar.
-   From $n = 100$ onwards all methods are close to nominal.

\newpage

## Power

```{r}
#| label: tbl-lin-power
metric_table(filter(lin, beta3 != 0), "power", "Power of the interaction test.")
```

```{r}
#| label: fig-lin-power
#| fig-cap: "Power."
metric_plot(filter(lin, beta3 != 0), "power", "Power")
```

-   Classical ML has the highest power, then reweighting, then MI; at small $n$ this ordering partly mirrors the type I error inflation.
-   For $\beta_3 = 0.035$ power stays low even at $n = 1000$; for $\beta_3 = 0.22$ all methods reach full power at $n = 1000$.

\newpage

## Coverage of $\beta_3$

```{r}
#| label: tbl-lin-cov
metric_table(lin, "coverage_beta3", "Coverage of the 95% Wald interval for beta3.")
```

```{r}
#| label: fig-lin-cov
#| fig-cap: "Coverage of the 95% interval for beta3 (dotted line: 0.95)."
metric_plot(lin, "coverage_beta3", "Coverage (beta3)", hline = 0.95)
```

-   Undercoverage at small $n$ for all methods; classical ML is worst, MI and reweighting are similar.
-   Coverage is close to nominal from $n = 50$ onwards.

\newpage

## Convergence

```{r}
#| label: tbl-lin-conv
convergence_table(lin, "Proportion of singular and not-converged replicates (pooled over effect sizes).")
```

```{r}
#| label: fig-lin-conv
#| fig-cap: "Convergence status, pooled over effect sizes."
#| fig-height: 5
convergence_plot(lin)
```

-   Singular fits occur at $n = 10$ for all methods at a similar rate and almost vanish from $n = 20$.
-   Reweighting and MI essentially always converge; classical ML has a small "not converged" share at every $n$, which grows sharply at $n = 1000$.
-   This is a false positive of lme4's absolute gradient check, which scales with the sample size: the flagged fits give the same estimates, coverage and power as the other fits and are kept in all metrics.

\newpage

## Computation time

```{r}
#| label: tbl-lin-time
metric_table(lin, "time_median_seconds", "Median computation time per replicate (seconds).")
```

```{r}
#| label: fig-lin-time
#| fig-cap: "Median computation time per replicate (log scale)."
metric_plot(lin, "time_median_seconds", "Median time (seconds)", log_y = TRUE)
```

-   Classical ML is fastest by far at every $n$.
-   MI is slowest up to $n = 100$; at $n = 1000$ MI and reweighting are similar.

\newpage

## Fraction of missing information (MI)

```{r}
#| label: tbl-lin-lambda
lambda_table(lin, "Mean fraction of missing information for beta3 (MI only).")
```

-   The fraction of missing information is higher under half-missing than under three observations minimum, and increases slightly with $n$.

\newpage

# Results - log time trend

-   MSE and coverage are not reported: the linear methods estimate a linear slope, which has no true value under the log trend.
-   Effect sizes: large ($\beta_1 = -2.42$, $\beta_3 = 1.95$), small ($\beta_1 = -0.385$, $\beta_3 = 0.31$).

## Type I error

```{r}
#| label: tbl-log-t1e
metric_table(filter(logt, beta3 == 0), "type1_error", "Type I error rate of the interaction test (identical arms).")
```

```{r}
#| label: fig-log-t1e
#| fig-cap: "Type I error rate (dotted line: nominal 0.05)."
metric_plot(filter(logt, beta3 == 0), "type1_error", "Type I error rate", hline = 0.05)
```

-   Small-$n$ inflation as in the linear scenarios; classical ML is again the most inflated.
-   Reweighting becomes conservative from $n = 50$ onwards and stays below nominal at $n = 1000$; classical ML and MI approach nominal.

\newpage

## Power

```{r}
#| label: tbl-log-power
metric_table(filter(logt, beta3 != 0), "power", "Power of the interaction test.")
```

```{r}
#| label: fig-log-power
#| fig-cap: "Power."
metric_plot(filter(logt, beta3 != 0), "power", "Power")
```

-   The linear methods still detect the crossing well: power is higher than in the linear scenarios with the same end-of-study difference (large vs. $\beta_3 = 0.22$, small vs. $\beta_3 = 0.035$).
-   Classical ML has the highest power; MI and reweighting are similar.

\newpage

## Convergence

```{r}
#| label: tbl-log-conv
convergence_table(logt, "Proportion of singular and not-converged replicates (pooled over effect sizes).")
```

```{r}
#| label: fig-log-conv
#| fig-cap: "Convergence status, pooled over effect sizes."
#| fig-height: 5
convergence_plot(logt)
```

-   Same pattern as the linear scenarios, with more singular fits at $n = 10$.
-   Classical ML's "not converged" share at $n = 1000$ is again the lme4 gradient false positive.

\newpage

## Computation time

```{r}
#| label: tbl-log-time
metric_table(logt, "time_median_seconds", "Median computation time per replicate (seconds).")
```

```{r}
#| label: fig-log-time
#| fig-cap: "Median computation time per replicate (log scale)."
metric_plot(logt, "time_median_seconds", "Median time (seconds)", log_y = TRUE)
```

-   Same ordering as the linear scenarios.

\newpage

## Fraction of missing information (MI)

```{r}
#| label: tbl-log-lambda
lambda_table(logt, "Mean fraction of missing information for beta3 (MI only).")
```

-   Slightly higher than in the linear scenarios; same pattern over dropout mechanism and $n$.

\newpage

# Suggested improvements

1.  LSPIM: make it scale beyond 50 clusters.
2.  Small-sample degrees of freedom: Satterthwaite for the classical ML test, Barnard-Rubin for the Rubin-pooled MI test; this targets the type I error inflation at small $n$.
3.  Classical ML convergence: replace lme4's absolute gradient flag with a relative gradient check, to remove the false "not converged" at large $n$.
