---
title: "Update Joint Modeling 07/09"
format: pdf
editor: visual
---

```{r}
#| label: setup
#| include: false

library(here)
library(ggplot2)
# Global chunk options
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

res <- readRDS(here::here("results", "data", "res.rds"))
res <- rbind(res[[1]]$summary, res[[2]]$summary, res[[3]]$summary)
res$n <- as.factor(res$n)
```

## Overview Methods:

The three suggested methods are implemented

1.  Classical ML

2.  MI + closed form

    1.  imputations were done within clusters, 3 imputations per clusters, with a maximum of 10 repetitions.

3.  Closed form + iterative reweighting:

    1.  **Currently reweighting until either max** $\Delta$ **of** $\beta$ **estimates \< 1e-6 or 30 iterations reached. Doing simple iterative reweighting: improvements are possible.**

    2.  **For the closed form, each cluster needs at least 3 observations in the current model (random intercept + random slope), current solution: exclude all clusters with too little observations. Perhaps better to only simulate data with at least**

\newpage

## Simulation setup

Formula: $$ y_{ij} = \beta_0 + \beta_1Treatment_i + \beta_2Time_j + \beta_3Treatment_i*Time_j + b_{0i} + b_{1i}*Time_j $$

Parameters were taken from the toenail data:

| Effect | Parameter | Size |
|------------------------|------------------------|------------------------|
| Intercept | $\beta_0$ | 2.4562 |
| Treatment Effect | $\beta_1$ | 0 *(0.593 in toenail data (this was a dose effect))* |
| Time Effect | $\beta_2$ | 0.2792 |
| Treatment x Time | $\beta_3$ | 0.0350 |
| Random intercept variance | $d_{11}$ | 7.3174 |
| Random slope variance | $d_{22}$ | 0.2239 |
| Random effects covariance | $d_{12}$ | -0.4985 |
| Residual variance | $\sigma^2$ | 3.1508 |

| Variable | Value |
|------------------------------------|------------------------------------|
| number of clusters | 100, 50, 20, 10 *(10.000 clusters was asked too, but currently excluded due to large computing time)* |
| number of repeated measures | 12 |
| missingness pattern | "half-missing" (half of the clusters have no missingness, other half have uniform probability to have 1-11 observations, dropout is monotone) |
| number of simulations / scenario | 10.000 |

## Next steps

1.  Improve convergence on reweighting

2.  Implement Type I error rate

3.  Joint model

## Results

### Mean Squared Error

```{r}
res[order(-as.numeric(res$n), res$method), c(2, 43, grep("^mse", names(res)))]
```

```{r}
ggplot(res, aes(x = n, y = mse_beta3, colour = method)) +
  geom_point() +
  labs(
    x = "Number of clusters",
    y = "MSE (beta3)",
    title = "MSE for beta3 by Method and Sample Size"
  ) +
  theme_minimal()
```


### Mean Absolute Bias

```{r}
res[order(-as.numeric(res$n), res$method), c(2, 43, grep("^mean_abs", names(res)))]
```

\newpage

### Convergence Rate

```{r}
res[order(-as.numeric(res$n), res$method), c(2, 43, grep("^prop_conv", names(res)))]
```

```{r}
ggplot(res, aes(x = n, y = (prop_converged_ok + prop_converged_warning), colour = method)) +
  geom_point() +
  labs(
    x = "Number of clusters",
    y = "Proportion of converged simulations",
    title = "Convergence Rate by Method and Sample Size"
  ) +
  theme_minimal()
```

\newpage

### Coverage rate of $\beta_3$

```{r}
res[order(-as.numeric(res$n), res$method), c(2, 43, grep("coverage95", names(res)))]
```

```{r}
ggplot(res, aes(x = n, y = coverage95_beta3, colour = method)) +
  geom_point() +
  labs(
    x = "Number of clusters",
    y = "Coverage rate of beta 3",
    title = "Coverage Rate of beta 3 in a 95%CI by Method and Sample Size"
  ) +
  geom_hline(yintercept = 0.95, linetype = "dotted") +
  theme_minimal()
```

\newpage

### Computation Time

```{r}
res[order(-as.numeric(res$n), res$method), c(2, 43, grep("seconds", names(res)))]
```

```{r}
ggplot(res, aes(x = n, y = time_median_seconds, colour = method)) +
  geom_point() +
  labs(
    x = "Number of clusters",
    y = "Median computation time (seconds)",
    title = "Median computation time by Method and Sample Size"
  ) +
  theme_minimal()
```