---
title: "Update Joint Modeling 09/07"
format: pdf
editor: visual
---

```{r}
#| label: setup
#| include: false

library(here)
# Global chunk options
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

res <- readRDS(here::here("results", "data", "agg_results.Rds"))
res <- rbind(res[[1]]$summary, res[[2]]$summary)
```

## Overview Methods:

Three methods for estimating model parameters were suggested:

1.  Classical maximum likelihood

2.  Multiple imputation + closed form

3.  Closed form + weighting

Currently:

1.  Classical ML completely implemented

2.  MI + closed form partly implemented: Fixed effects and variance is implemented, RE and variance estimators of fixed effects not yet implemented **Q: current quick fix solution by using equal weights, is this ok?**

3.  Closed form + weighting: not yet implemented

## Preliminary Results

### Simulation setup

Formula: $$ y_{ij} = \beta_0 + \beta_1Treatment_i + \beta_2Time_j + \beta_3Treatment_i*Time_j + b_{0i} + b_{1i}*Time_j $$

Parameters were taken from the toenail data:

| Effect | Parameter | Size |
|----|----|----|
| Intercept | $\beta_0$ | 2.4562 |
| Treatment Effect | $\beta_1$ | 0 **0.593 in toenail data (this was a dose effect)** |
| Time Effect | $\beta_2$ | 0.2792 |
| Treatment x Time | $\beta_3$ | 0.0350 |
| Random intercept variance | $d_{11}$ | 7.3174 |
| Random slope variance | $d_{22}$ | 0.2239 |
| Random effects covariance | $d_{12}$ | -0.4985 |
| Residual variance | $\sigma^2$ | 3.1508 |

| Variable | Value |
|------------------------------------|------------------------------------|
| number of clusters | 100, 50, 20, 10 *(10.000 clusters was asked too, but currently excluded due to large computing time)* |
| number of repeated measures | 12 |
| missingness pattern | "half-missing" (half of the clusters have no missingness, other half have uniform probability to have 1-11 observations, dropout is monotone) |
| number of simulations / scenario | 100 *(will use 1000 in the future, but restricted due to computing time)* |

For the imputations: imputations were done within clusters, 3 imputations per clusters, with a maximum of 10 repetitions.

## Results

### Mean Absolute Bias

Larger bias in classical ml

```{r}
res[order(-res$n), c(2, 33, grep("^mean_abs", names(res)))]
```

### Mean Relative Bias

Note: probably better to use mean absolute relative bias?

```{r}
res[order(-res$n), c(2, 33, grep("^mean_rel", names(res)))]
```

\newpage

### Convergence Rate

Rate of convergence not yet implemented for mi + closed form. Convergence for classical ML needs checking

```{r}
res[order(-res$n), c(2, 33, grep("^prop_conv", names(res)))]
```

### Coverage rate of $\beta_3$

Coverage rate for MI + closed form not yet implemented (no variance of estimates yet)

```{r}
res[order(-res$n), c(2, 33, grep("coverage95", names(res)))]
```

### Computation Time

Classical ML is faster (due to time required for imputation)

```{r}
res[order(-res$n), c(2, 33, grep("seconds", names(res)))]
```

## Next steps

1.  Check rate of convergence

2.  Implement the calculation of RE variances and FE estimate variances for MI + closed form

3.  Implement closed form + weighting

4.  Present results with graphs

5.  Joint model
